---
title: "Testing parameters for facilitation and competition networks"
---


Script to try parameters out and explore the output.

I haven't found empirical network data for facilitation and competition so far. The evaluation I am doing consist mainly in exploring the species richness curve (expecting a progressive increase with time and then an estabilization), degree distribution (expecting diversity of values), and some model objects (trait evolution, speciation/extinction events).






```{r}
library(EnvStats)
library(foreach)
library(ggplot2)
library(igraph)
library(ggpubr)
library(reshape2)
library(tidyr)
library(dplyr)

```


For plotting:

```{r}
my.theme<-theme(axis.text=element_text(size=12),
                axis.title = element_text(size = 14),
                legend.position = "top",
                legend.text=element_text(size=10),
                legend.title = element_text(size=12),
                plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
                axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0)),
                axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)))


plot_spp_richness <- function(steps, pres_mat, title) {
  
  timesteps <- 1:steps
  n_spp_time <- c()
  
  for (i in 1:nrow(pres_mat)) {
  
  n_spp_time[i] <- length(which(pres_mat[i,] == 1))
  
  }
  
  df_spprich_time <- data.frame(timesteps,n_spp_time)
  
  p.spp <- ggplot(df_spprich_time, aes(x=timesteps, y = n_spp_time)) +
  geom_line(color="black", linetype="twodash") +
  theme_classic()+
  my.theme+
  xlab("timesteps")+
  ylab("N species")+
    ggtitle(title)
  
  return(p.spp)
}

```



```{r}
source("../functions/functions_simulation_fac_comp.R")
source("../functions/function_plot_mean_div.R")
```



```{r}

pars = list()

# General parameters
#pars$int = 0 # 0 for competition, 1 for facilitation, 2 for predation
pars$Smax = 1000 # Maximal number of species in the system

# Related to interaction range trait
pars$av_r = 0.1 # Half range of the niche of the first species
pars$sd = 0.5*pars$av_r + 0.0001 # Standard deviation of the normal distribution used to calculate the niche optimum trait of a new species

pars$u_max = 0.15 # Speciation probability

# Extinction probability, negative interactions communities
pars$a_eneg = 0.025 # Shape of the exponential decay of the negative extinction - interaction relationship
pars$e_0neg = 0.5 # Asymptotic extinction probability with infinite negative interactions
pars$e_1neg = 1  # Extinction probability with absence of interactions

# Extinction probability, positive interactions communities
pars$a_epos = 1.2  # Shape of the exponential decay of the positive extinction - interaction relationship
pars$e_0pos = 0.1 #0.075 # Asymptotic extinction probability with infinite positive interactions
pars$e_1pos = 4.8 #5.19 # 1 - pars$e_0pos

# Establishment probability, negative interactions communities
pars$a_uneg = 0.075 # Shape of the exponential decay of the colonization - interaction relationship
pars$u_0neg = 0.075 # Asymptotic establishment probability with infinite competition interactions
pars$u_1neg = 2 # Establishment probability with absence of competition interaction

# Establishment probability, positive interactions communities
pars$a_u = 0.8 #0.45 # Shape of the exponential decay of the colonization - interaction relationship
pars$u_0 = 1 # Asymptotic establishment probability with infinite facilitation interactions
pars$u_1 = -0.9 #-1 # Establishment probability with absence of facilitation interactions
pars$d = 0.5 # Decrease speed of the establishment probability

# Extinction and establishment probabilitie for both, positive and negative interactions communities
pars$Bspe = 4 # Constant minimal number of interaction per species (extablishment prob)
pars$Bext = 4 # Constant minimal number of interaction per species (extinction prob)
pars$I_max = 60 # Maximal number of interactioning species


pars$beta_n = 1 # parameter of the beta distribution


# STRENGTH OF NEUTRAL-DRIVEN EVOLUTION

pars$SN = 1


# PROBABILITY OF ESTABLISHMENT
# strength of selection-driven selection is 1 - SN
pars$estab_prob_neutral = 0.5 

# PROBABILITY OF EXTINCTION

pars$ext_prob_neutral = 0.03 # neutral probability of extinction

```




tests


```{r}
# Extinction probability parameters
pars$e_0pos <- 0.1  # Reduce the asymptotic extinction probability with infinite positive interactions
pars$e_1pos <- 4.8  # Slightly decrease the extinction probability in the absence of positive interactions
pars$a_epos <- 1.2  # Decrease the rate at which extinction probability decreases with more positive interactions

# Establishment probability parameters
pars$u_0 <- 1.0  # Increase the asymptotic establishment probability with infinite positive interactions
pars$u_1 <- -0.9 #-0.9  # Increase the establishment probability in the absence of positive interactions
pars$a_u <- 0.8  # Keep the rate at which establishment probability decreases with more positive interactions


```



# set parameters simulation

```{r}
nsteps = 250 # Set the maximum timestep per simulation
nsim = 2 # Set the number of simulations

# Vector to record number of simulation needed to reach 100 valid simulations
needed_sim_pos <- c(rep(NA, nsim))
needed_sim_neg <- c(rep(NA, nsim))
```


# Simulation




```{r}
# Initialize lists outside the loop
list_results_fac <- list()
list_results_comp <- list()

# Loop through interaction types (0 and 1)
for(int in 0:1) {
  pars$int <- int
  
  if(int == 1) {
    print("Positive interactions")
  } else {
    print("Negative interactions")
  }
  
  successful_sim <- 1  # Initialize count of successful simulations
  total_nb_sim <- 0    # Initialize total count of simulations
  seed_list <- integer(0)  # Initialize seed list
  
  while(successful_sim <= nsim) {
    
    seed <- sample.int(80000, 1)  # Generate a random seed
    
    # Check if seed has already been used
    while(seed %in% seed_list) {
      seed <- sample.int(80000, 1)
    }
    
    seed_list <- c(seed_list, seed)  # Add seed to the list
    
    # Run simulation
    simulation <- sim_model_bif_fc(seed = seed, pars, nsteps = nsteps)
    
    total_nb_sim <- total_nb_sim + 1  # Increment total simulations
    
    # Check if simulation has enough species at timestep 150
    success <- sum(simulation$pres[150, ])
    
    if(success >= 20) {
      # Record results from the simulation
      res_sim <- list(simulation_number = paste0("simulation", successful_sim),
                      seed = seed,
                      parameters = pars,
                      presence_matrix = simulation$pres,
                      traits_df = simulation$traits,
                      parentage_matrix = simulation$anc,
                      extinction_matrix = simulation$extinct,
                      network_list = simulation$L_list,
                      anc_dist_table = simulation$dist_anc,
                      list_anc_dist = simulation$list_dist_anc)
      
      if(int == 1) {
        list_results_fac[[successful_sim]] <- res_sim
      } else if(int == 0) {
        list_results_comp[[successful_sim]] <- res_sim
      }
      
      print(paste("Simulation", successful_sim, "of", nsim, "successful out of", total_nb_sim))
      successful_sim <- successful_sim + 1  # Increment successful simulation count
    }
  }
}

```



## Speciation and extinction events - mean among simulations


```{r}
mean_div_curves_fac <- compute_div_curves_from_list_res(list_results_fac)[[1]]
mean_div_curves_comp <- compute_div_curves_from_list_res(list_results_comp)[[1]]

ggarrange(mean_div_curves_fac + labs(title ="facilitation"),
           mean_div_curves_comp + ylab(NULL) + labs(title = "competition"),
           nrow = 1,
           ncol = 2, 
          common.legend = TRUE)

```



## Inspection of individual simulation


```{r}

inspect_simulation_fac_comp(simulation_data = list_results_fac[[2]], interaction_type = "facilitation")

inspect_simulation_fac_comp(simulation_data = list_results_comp[[2]], interaction_type = "competition")


```






