---
title: "R Notebook"
---

```{r}
library(phytools)

source("functions/functions_phylosignal.R")
library(ape)
library(apTreeshape)
library(aricode)
library(igraph)
library(ggplot2)
library("vegan")
library(ggpubr)
library(ade4)
```


Example Phytools

```{r}
## load data from Garland et al. (1992)
data(mammal.tree)
data(mammal.data)
## extract characters of interest
ln.bodyMass<-log(setNames(mammal.data$bodyMass,
rownames(mammal.data)))
ln.homeRange<-log(setNames(mammal.data$homeRange,
rownames(mammal.data)))


###############################
## compute phylogenetic signal K
###############################

K.bodyMass<-phylosig(mammal.tree,ln.bodyMass,
test=TRUE)
print(K.bodyMass)
plot(K.bodyMass)
K.homeRange<-phylosig(mammal.tree,ln.homeRange,
test=TRUE)
print(K.homeRange)
plot(K.homeRange)

################################
## compute phylogenetic signal lambda
################################

lambda.bodyMass<-phylosig(mammal.tree,ln.bodyMass,
method="lambda",test=TRUE)
print(lambda.bodyMass)
plot(lambda.bodyMass)
lambda.homeRange<-phylosig(mammal.tree,ln.homeRange,
method="lambda",test=TRUE)
print(lambda.homeRange)
plot(lambda.homeRange)

```



### Load simulation data

```{r}
Smax = 1000
```


FACILITATION

```{r}

list_res <- readRDS("../Data/selection_evolution/facilitation/test/list_res_pos_2.rds")


# count number of timesteps where there are spp

list_simulation1 <- list_res

n_steps <- length(list_simulation1$network_list)


```


#### Crop presence matrix to nsteps

The presence_matrix (rows = timesteps, cols = species, entries = 1,0) has extra timesteps in case the simulation is longer (max = 350). Crop the presence matrix to the number of timesteps in the simulation:

```{r}

presence_matrix <- list_simulation1$presence_matrix

presence_matrix <- presence_matrix[1:n_steps,]

```


#### Identify timesteps where phylogenetic distances cant be calculated

Im going to discard those with 1 spp

```{r}

non.valid_timesteps_phylo_distance <- c(which(rowSums(presence_matrix) < 2))

# until what timestep need to discard:

final.discarded_timestep <- non.valid_timesteps_phylo_distance[length(non.valid_timesteps_phylo_distance)]

print(paste("timesteps 1 to", final.discarded_timestep, "will be discarted as they have less than 3 spp"))


```


#### homogenize elements to start from valid timesteps

Discard those timesteps with less than 2 spp in both the list_anc_dist table (ancestry table) and the network_list:

```{r}

# ancestry-distances table

list_anc_dist <- list_simulation1$list_anc_dist[(final.discarded_timestep+1):length(list_simulation1$list_anc_dist)]



# Network list

network_list <- list_simulation1$network_list[(final.discarded_timestep+1):length(list_simulation1$list_anc_dist)]







## ------------------- Check

if (length(which(is.null(list_simulation1$network_list))) > 0) {
  
  print("check - null network somewhere")
} else {
  print("all good - no null networks")
}


if(length(list_anc_dist) == length(network_list)){
  
  print("all good - distance-ancestry table and network list have the same length")
} else {
  
  print("PROBLEM")
}


```



#### Convert spp names from numbers to letters


Species names in both the ancestry table and the networks are numbers. Convert it into letters with the function **change_sppnames_letters_ancdist.table()** for the ancestry table, and **set_sppNames_numbers()** (because matrices lack colnames and rownames) and **convert_sppnames_toletters()** for the network list:

```{r}

## ancestry-distances table

list_anc_dist_letters <- lapply(list_anc_dist, change_sppnames_letters_ancdist.table)


## Network list 


list_networks_sppnames_numbers <- lapply(network_list, set_sppNames_numbers)

#### convert numbers to letters

list_networks_sppnames_letters <- lapply(list_networks_sppnames_numbers, convert_sppnames_toletters)






## ------------------- Check


if(length(list_anc_dist_letters) == length(list_networks_sppnames_letters)){
  
  print("all good - distance-ancestry table with letters and network list with leters have the same length")
} else {
  
  print("PROBLEM")
}


```


#### convert spp names to letters in presence matrix



```{r}

# Set the same spp names for the presence_matrix than for the interacion matrices


colnames(presence_matrix) <- seq(1:Smax)

colnames(presence_matrix) <- chartr("0123456789", "ABCDEFGHIJ", colnames(presence_matrix))



## ------------------- Check

if(length(which(colnames(presence_matrix) != colnames(list_networks_sppnames_letters[[1]]))) == 0){
  
  print("all good - spp names in presence_matrix correspond to spp names in network list")
  
} else {
  
  print("PROBLEM")
} 





# Discard same timesteps (rows) than the discarted phylogenetic distance matrices

presence_matrix <- presence_matrix[(final.discarded_timestep+1):length(list_simulation1$list_anc_dist),]




## ------------------- Check

if(nrow(presence_matrix) == length(list_networks_sppnames_letters) & 
   nrow(presence_matrix) == length(list_dist.phylo)){
  
  print("all good - presence matrix, list phylo dist and list interaction networks have the same n_steps")
  
} else {
  
  print("PROBLEM")
}




```




## compute trees

```{r}
list_newick <- list()
list_trees <- list()
list_newick_tails <- list()
list_dist.phylo <- list()



for (i in 1:length(list_anc_dist_letters)) {
  
  
  #print(sprintf("step %s", i))
  
  list_newick[[i]] <- ToPhylo(list_anc_dist_letters[[i]])
  
  list_newick_tails[[i]] <- paste(list_newick[[i]], "root")
  
  list_trees[[i]] <- read.tree(text = sub("A root",";",list_newick_tails[[i]]))
  
  list_dist.phylo[[i]] <- cophenetic.phylo(list_trees[[i]])
  

}


list_anc_dist_letters[[1]]
plot(list_trees[[20]])

length(list_trees[[20]]$tip.label) == length(which(presence_matrix[20,] == 1))



```





