---
title: "Testing parameters for Foodweb"
---

```{r}
library(ggplot2)
library(ggpubr)
library(reshape2)
library(igraph)
library(dplyr)
```


Load functions

```{r}
source("../functions/functions_metrics_networks.R")
```

Parameters for plotting

```{r}
my.theme<-theme(axis.text=element_text(size=12),
                axis.title = element_text(size = 14),
                legend.position = "top",
                legend.text=element_text(size=10),
                legend.title = element_text(size=12),
                plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
                axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0)),
                axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)))





```


Metrics of empirical foodwebs:

empirical networks data was obtained from the ECOWeB 1.1 database:
https://digitalcommons.rockefeller.edu/cohen_joel_laboratory/1/

```{r}
net_metr_empirical <- readRDS(here::here("data/net_metrics_empirical.rds"))

net_metr_empirical$id <- NULL

# eliminate one problematic foodweb:

net_metr_empirical <- net_metr_empirical[-13,]

# reshape dataframe:

df_metrics_emp <- reshape_df_net.metrics(df_metrics = net_metr_empirical,
                       interaction = "empirical")
```




```{r}
source("../functions/functions_simulation_fw.R")
source("../functions/functions_metrics_networks.R")
source("../functions/function_plot_mean_div.R")
source(here::here("code/functions/functions_simulation_inspection.R"))
```


[Notes]

- At range 0.18, I obtain a rich diversity of n values and top predators, but interestingly, the community collapses twice into a specific range of niche values.



```{r}

pars = list()



pars$competition_coefficient = 0.025

pars$Sbasal = 25 # number of basal species
pars$Smax = 1000 # Maximal number of species allowed

pars$Bspe = 4 # Basal species impact on the speciation probality
pars$Bext = 4 # Basal species impact on extinction probability

pars$av_r = 0.185 # 0.2 range of the niche
#pars$sd = 0.5*pars$av_r + 0.0001 # Standard deviation of the normal distribution used to calculate the niche optimum trait of a new species


# PROBABILITY OF MUTATION

pars$u_max = 0.23 #0.15 # mutation probability (0.075, )
pars$d = 0.5 # Decrease speed of the establishment probability
pars$I_max = 60 # Maximal number of interactioning species

pars$beta_n =  1 # parameter of the beta distribution 



# STRENGTH OF NEUTRAL-DRIVEN EVOLUTION

pars$SN = 0 # strength for neutral-driven evolution



# PROBABILITY OF ESTABLISHMENT


# strength of selection-driven selection is 1 - SN
pars$estab_prob_neutral = rep(0, pars$Smax) # neutral probability of establishment



# Facilitation & Foodweb

pars$u_0pos = 1  
pars$u_1pos = -1 
pars$a_upos = 0.45 



# PROBABILITY OF EXTINCTION

pars$ext_prob_neutral = 0 # neutral probability of extinction


# Competition

pars$e_0neg = 0.1 #0.15 # Asymptotic extinction probability with infinite negative interactions
pars$a_eneg = 0.025 # Shape of the exponential decay of the negative extinction - interaction relationship
pars$e_1neg = -pars$e_0neg  # Extinction probability with absence of interactions

# Facilitation & Foodweb

pars$e_0pos = 0.075 
pars$e_1pos = 5.19 
pars$a_epos = 1.2 



#saveRDS(pars, "pars_fw_12june.rds")

#pars <- readRDS("pars_work_12june.rds")

```


# set parameters simulation

```{r}
nsteps = 250 # Set the maximum timestep per simulation
nsim = 2 # Set the number of simulations
```




### Test one interaction


```{r}


pars$int = 2

list_results <- list()
#list_adj_matrix.f <- list()

for (i in 1:nsim) {
  
  
  seed <- set.seed(50)
  
  simulation <- sim_model(seed = runif(1), 
                        pars, 
                        nsteps = nsteps)
  
  res <- list(
                      seed = seed,
                      parameters = pars, 
                     presence_matrix = simulation$pres, 
                      traits_df = simulation$traits,
                      parentage_matrix = simulation$anc,
                      extinxction_matrix = simulation$extinct, 
                      network_list = simulation$L_list,
                       anc_dist_table = simulation$dist_anc, 
                       list_anc_dist = simulation$list_dist_anc
      ) # Record results from the simulation
  
  list_results[[i]] <- res
  
     
 # list_adj_matrix.f[[i]] <- get_adjancency_from_simulation.fw_single(list_networks = res$network_list,
                                       #    nbasals = pars$Sbasal,
                                       #    pres_mat = res$presence_matrix)
  
}



#saveRDS(list_results, "list_results_fw_10sim.rds")

```



## Speciation and extinction events - mean among simulations


```{r}
compute_div_curves_from_list_res(list_results)[[1]]
```


## Inspection of individual simulation

```{r}
inspect_simulation_fw(simulation_data = list_results[[1]])
```




## Further checks for foodweb



```{r}

list_adj_matrix.f <- list()

for (i in 1:length(list_results)) {
  
  
  res <- list_results[[i]]
  
  list_adj_matrix.f[[i]] <- get_adjancency_from_simulation.fw_single(list_networks = res$network_list,
                                           nbasals = pars$Sbasal,
                                           pres_mat = res$presence_matrix)
  
}

net_metr_sim <- get_network_measures_from_adjacency(list_adj_matrix = list_adj_matrix.f)


df_metrics_fw <- reshape_df_net.metrics(df_metrics = net_metr_sim,
                       interaction = "foodweb")

df_metrics <- merge(df_metrics_fw,
                    df_metrics_emp,
                    by = "metric")


# plot the results

p_S <- plot_metric_foodweb(df = df_metrics, metric = "S", y_axis = "N species")
p_link.dens <- plot_metric_foodweb(df = df_metrics, metric = "Link_density", y_axis = "L/S")
p_C <- plot_metric_foodweb(df = df_metrics, metric = "C", y_axis = "Connectance")
p_top <- plot_metric_foodweb(df = df_metrics, metric = "perc_tops", y_axis = "% top species")
p_int <- plot_metric_foodweb(df = df_metrics, metric = "perc_int", y_axis = "% intermediate species")
p_basal <- plot_metric_foodweb(df = df_metrics, metric = "perc_basals", y_axis = "% basal species")
p_cannibals <- plot_metric_foodweb(df = df_metrics, metric = "perc_cannibals", y_axis = "% cannibal species")
p_omnivory <- plot_metric_foodweb(df = df_metrics, metric = "omnivory", y_axis = "omnivory")
p_sd_gen <- plot_metric_foodweb(df = df_metrics, metric = "sd_gen", y_axis = "SD generality")
p_sd_vul <- plot_metric_foodweb(df = df_metrics, metric = "sd_vul", y_axis = "SD vulnerability")

plot_metrics_comparison_sim_emp <- ggarrange(
  p_S,
  p_link.dens,
  p_C,
  p_top,
  p_int,
  p_basal,
  p_cannibals,
  p_omnivory,
  p_sd_gen,
  p_sd_vul,
  
  nrow = 5,
  ncol = 2
  
)

plot_metrics_comparison_sim_emp

#ggsave("plot_metrics_comparison_sim_emp.png", height = 12, width = 7)
```




