---
title: "R Notebook"

---

```{r}
library(dplyr)
library(tidyr)
```


### get paths simulations results

Read the results of n simulations produced by simulation_fac_comp.Rmd, simulation_foodweb.Rmd, and simulation_neutral.Rmd scripts:

```{r}

path_sim_fac <-"C:/Users/alexf/Desktop/PhD_Fuster-Calvo/Chapter 2 - Network_phylogeny/Network-Phylogeny/data/selection_evolution/facilitation/test"
file_paths_fac <- list.files(path = path_sim_fac, pattern = ".rds", full.names = TRUE)


list_res <- readRDS(file_paths_fac[1])
```




```{r}
# count number of timesteps where there were spp

n_steps <- length(list_res$network_list)

network_list <- list_res$network_list

presence_matrix <- list_res$presence_matrix[1:n_steps,]
speciation_matrix <- list_res$speciation_matrix[1:n_steps,]
extinction_matrix <- list_res$extinction_matrix[1:n_steps,]


null_indices <- sapply(network_list, is.null)


if(length(which(null_indices)) !=0){
  
  network_list <- network_list[-which(null_indices)]
  
  presence_matrix <- presence_matrix[-which(null_indices),]
  speciation_matrix <- speciation_matrix[-which(null_indices),]
  extinction_matrix <- extinction_matrix[-which(null_indices),]
}



# Set the same spp names for the presence_matrix than for the interacion matrices


colnames(presence_matrix) <- seq(1:1000)
colnames(speciation_matrix) <- seq(1:1000)
colnames(extinction_matrix) <- seq(1:1000)

colnames(presence_matrix) <- chartr("0123456789", "ABCDEFGHIJ", colnames(presence_matrix))
colnames(speciation_matrix) <- chartr("0123456789", "ABCDEFGHIJ", colnames(speciation_matrix))
colnames(extinction_matrix) <- chartr("0123456789", "ABCDEFGHIJ", colnames(extinction_matrix))


```



```{r}
# Function to change column names
change_colnames <- function(mat) {
  colnames(mat) <- chartr("0123456789", "ABCDEFGHIJ", colnames(mat))
  return(mat)
}



# Assign column names to matrices (if not already assigned)
for (mat in 1:length(network_list)) {
  
  colnames(network_list[[mat]]) <- seq_len(ncol(network_list[[mat]]))  # Assigning default column names
}

# Apply the function to each matrix in the list
network_list <- lapply(network_list, change_colnames)

```


In the list_anc_dist I have in each timestep the code of species (number) and whether it went extinct and the ancestor species. From this I can get when it went extinct (first time that E appears for it), the speciation events, when it appears as an ancestor for every new species.

In the network, I can crop it and compute centrality. But I need to make sure I keep the numbers of species to match the ones I get from extinction ans speciation information. I could convert them into letters first, then crop the network, then compute centrality.




### Add letter IDs to species in speciation and extinction matrices

```{r}

df_speciation <- as.data.frame(speciation_matrix)
df_extinction <- as.data.frame(extinction_matrix)

df_speciation$timestep = seq(from = 1, to = nrow(df_speciation), 1)
df_extinction$timestep = seq(from = 1, to = nrow(df_extinction), 1)

```



```{r}

#from functions_phylosignal.R

set_sppNames_numbers <- function(mat) {
  
  rownames(mat) <- seq(1:nrow(mat))
  colnames(mat) <- seq(1:ncol(mat))
  
  return(mat)  
}

convert_sppnames_toletters <-  function(mat) {
  
  
  rownames(mat) <- chartr("0123456789", "ABCDEFGHIJ", rownames(mat))
  colnames(mat) <- chartr("0123456789", "ABCDEFGHIJ", colnames(mat))
  
  return(mat)
  
}

```




### spp names in networks + crop networks + obtain centrality per letter ID

```{r}

    ## Network list 
    

list_networks_sppnames_numbers <- lapply(network_list, set_sppNames_numbers)


#### convert numbers to letters

list_networks_sppnames_letters <- lapply(list_networks_sppnames_numbers, convert_sppnames_toletters)



for (i in 1:length(list_networks_sppnames_letters)) {
  
  
  list_net_present_spp.letters[[i]] <- list_networks_sppnames_letters[[i]][names(which(presence_matrix[i,] == 1)), names(which(presence_matrix[i,] == 1))]
  
}



# crop the speciation and extinction matrices by the present species

list_speciation <- list()
list_extinction <- list()

########################################################### PROBLEM HERE - For some reason, the speciation or extinction matrices are one timestep ahead of networks, meaning that list_network[20] corresponds to speciation_matrix[21,]. I correct it below with the i+1:


for(i in 1:(length(list_net_present_spp.letters)-1)){
  
  list_speciation[[i]] <- speciation_matrix[i+1,colnames(speciation_matrix) %in% colnames(list_net_present_spp.letters[[i]])]
  list_extinction[[i]] <- extinction_matrix[i+1,colnames(extinction_matrix) %in% colnames(list_net_present_spp.letters[[i]])]
}



# Transform into long format


list_speciation_df <- lapply(list_speciation, function(x) {
  df <- data.frame(species = names(x), speciation = as.numeric(x))
  return(df)
})

list_extinction_df <- lapply(list_extinction, function(x) {
  df <- data.frame(species = names(x), extinction = as.numeric(x))
  return(df)
})



# Add timestep to each dataframe and bind them


list_speciation_with_timestep <- lapply(seq_along(list_speciation_df), function(i) {
  df <- list_speciation_df[[i]]
  df$timestep <- i
  return(df)
})

# Add timestep to each dataframe and bind them
list_extinction_with_timestep <- lapply(seq_along(list_extinction_df), function(i) {
  df <- list_extinction_df[[i]]
  df$timestep <- i
  return(df)
})


```

(note that timestep number is not correct. I need to take into account the null_indices - I can do this directly changing it in the df for the plot)

```{r}
# Bind rows of all dataframes in the list

df_all_speciation <- do.call(rbind, list_speciation_with_timestep)
df_all_extinction <- do.call(rbind, list_extinction_with_timestep)

```



#########

Now the normalized degree per species and timestep

```{r}
compute_degree <- function(network) {
  # Sum the connections of each species
  degrees <- rowSums(network)
  
  # Check if the degrees vector is empty
  if (length(degrees) == 0 || all(degrees == 0)) {
    df <- data.frame(species = character(), degree = numeric())
  } else {
    # Normalize the degrees
    normalized_degrees <- degrees / sum(degrees)
    
    # Create a dataframe
    df <- data.frame(species = names(degrees), norm_degree = normalized_degrees, degree = degrees)
  }
  
  return(df)
}



# Compute normalized degree for each network
list_normalized_degree <- lapply(list_net_present_spp.letters, compute_degree)


# Add timestep

list_normalized_degree_timestep <- lapply(seq_along(list_normalized_degree), function(i) {
  df <- list_normalized_degree[[i]]
  df$timestep <- i
  return(df)
})

# Bind rows of all dataframes in the list

df_all_degree <- do.call(rbind, list_normalized_degree_timestep)


```

Bind dataframes

```{r}


merged_df <- merge(df_all_degree, df_all_speciation, by = c("species", "timestep"))
merged_df <- merge(merged_df, df_all_extinction, by = c("species", "timestep"))


```

Plot

```{r}
library(ggplot2)

# Convert speciation to factor for coloring
merged_df$speciation <- factor(merged_df$speciation)

# Create the plot
ggplot(merged_df, aes(x = timestep, y = norm_degree, color = speciation)) +
  geom_point(size = 3, alpha = 0.4) +
  scale_color_manual(values = c("black", "red", "darkred"),
                     labels = c("Extinction", "Speciation (1)", "Speciation (2)")) +
  labs(x = "Timestep", y = "Degree", color = "Event") +
  theme_minimal()

# Create the plot
ggplot(merged_df, aes(x = timestep, y = degree, color = speciation)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("black", "red", "darkred"),
                     labels = c("Extinction", "Speciation (1)", "Speciation (2)")) +
  labs(x = "Timestep", y = "Degree", color = "Event") +
  theme_minimal()

```

```{r}
# Filter dataframe for extinctions and speciations
extinction_df <- merged_df[merged_df$extinction == 1, ]
speciation_df <- merged_df[merged_df$speciation != 0, ]

# Create the plot
ggplot() +
  geom_line(data = extinction_df, aes(x = timestep, y = norm_degree, group = 1), color = "black") +
  geom_line(data = speciation_df, aes(x = timestep, y = norm_degree, group = 1), color = "red", size = 1.5) +
  labs(x = "Timestep", y = "Degree") +
  theme_minimal()

# Create the plot
ggplot() +
  geom_line(data = extinction_df, aes(x = timestep, y = degree, group = 1), color = "black") +
  geom_line(data = speciation_df, aes(x = timestep, y = degree, group = 1), color = "red", size = 1.5) +
  labs(x = "Timestep", y = "Degree") +
  theme_minimal()

```

```{r}
# Calculate mean degree for each timestep
extinction_mean <- aggregate(degree ~ timestep, data = extinction_df, FUN = mean)
speciation_mean <- aggregate(degree ~ timestep, data = speciation_df, FUN = mean)

# Create the plot
ggplot() +
  geom_line(data = extinction_mean, aes(x = timestep, y = norm_degree), color = "black") +
  geom_line(data = speciation_mean, aes(x = timestep, y = norm_degree), color = "red", size = 1.5) +
  labs(x = "Timestep", y = "Mean Degree", color = "Event") +
  theme_minimal()


# Create the plot
ggplot() +
  geom_line(data = extinction_mean, aes(x = timestep, y = degree), color = "black") +
  geom_line(data = speciation_mean, aes(x = timestep, y = degree), color = "red", size = 1.5) +
  labs(x = "Timestep", y = "Mean Degree", color = "Event") +
  theme_minimal()

```




```{r}
# Function to categorize degrees
categorize_degree <- function(degree) {
  breaks <- quantile(degree, probs = c(0, 1/3, 2/3, 1))
  cut(degree, breaks = breaks, labels = c("low", "medium", "high"), include.lowest = TRUE)
}

# Categorize degree
merged_df1 <- merged_df %>%
  mutate(degree_category = categorize_degree(degree))

# Count the number of extinctions, speciations, and total species for each category
summary_df_degree <- merged_df1 %>%
  group_by(degree_category) %>%
  summarise(total_species = n(),
            extinction_count = sum(extinction),
            speciation_count = sum(speciation != 0))

# Reshape data for plotting
summary_df_degree <- tidyr::pivot_longer(summary_df_degree, cols = c(extinction_count, speciation_count, total_species),
                                  names_to = "event", values_to = "count")

# Create the plot for degree
ggplot(summary_df_degree, aes(x = degree_category, y = count, fill = event)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Degree Category", y = "Count", fill = "Event") +
  scale_fill_manual(values = c("black", "red", "blue"), labels = c("Extinction", "Speciation", "Total Species")) +
  theme_minimal()

# Function to categorize normalized degrees
categorize_norm_degree <- function(norm_degree) {
  breaks <- quantile(norm_degree, probs = c(0, 1/3, 2/3, 1))
  cut(norm_degree, breaks = breaks, labels = c("low", "medium", "high"), include.lowest = TRUE)
}

# Categorize normalized degree
merged_df1 <- merged_df1 %>%
  mutate(norm_degree_category = categorize_norm_degree(norm_degree))

# Count the number of extinctions, speciations, and total species for each category
summary_df_norm_degree <- merged_df1 %>%
  group_by(norm_degree_category) %>%
  summarise(total_species = n(),
            extinction_count = sum(extinction),
            speciation_count = sum(speciation != 0))

# Reshape data for plotting
summary_df_norm_degree <- tidyr::pivot_longer(summary_df_norm_degree, cols = c(extinction_count, speciation_count, total_species),
                                  names_to = "event", values_to = "count")

# Create the plot for normalized degree
ggplot(summary_df_norm_degree, aes(x = norm_degree_category, y = count, fill = event)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Normalized Degree Category", y = "Count", fill = "Event") +
  scale_fill_manual(values = c("black", "red", "blue"), labels = c("Extinction", "Speciation", "Total Species")) +
  theme_minimal()
```

