---
title: "Understanding parameters"

---


# Probability of extinction


 ext_prob_sel <- e_0neg + e_1neg * exp(-a_eneg * avg_similarity)


```{r}
# Load necessary libraries
library(ggplot2)
library(gridExtra)

# Define a range of average similarity values
avg_similarity <- seq(0, 1, length.out = 100)

# Define different sets of parameters to explore their effects
e_0neg_values <- c(0.1, 0.2, 0.3)  # Base extinction probabilities
e_1neg_values <- c(-0.1, 0.1, 0.2)  # Modifiers for extinction probability
a_eneg_values <- c(0.025, 0.1, 0.2)  # Decay rates for exponential term

# Create data for different values of e_0neg
results_e0neg <- data.frame()
for (e_0neg in e_0neg_values) {
  ext_prob_sel <- e_0neg + (-e_0neg) * exp(-0.025 * avg_similarity)
  results_e0neg <- rbind(results_e0neg, data.frame(avg_similarity = avg_similarity, ext_prob = ext_prob_sel, parameter = paste("e_0neg =", e_0neg)))
}

# Create data for different values of e_1neg
results_e1neg <- data.frame()
for (e_1neg in e_1neg_values) {
  ext_prob_sel <- 0.1 + e_1neg * exp(-0.025 * avg_similarity)
  results_e1neg <- rbind(results_e1neg, data.frame(avg_similarity = avg_similarity, ext_prob = ext_prob_sel, parameter = paste("e_1neg =", e_1neg)))
}

# Create data for different values of a_eneg
results_aeneg <- data.frame()
for (a_eneg in a_eneg_values) {
  ext_prob_sel <- 0.1 + (-0.1) * exp(-a_eneg * avg_similarity)
  results_aeneg <- rbind(results_aeneg, data.frame(avg_similarity = avg_similarity, ext_prob = ext_prob_sel, parameter = paste("a_eneg =", a_eneg)))
}

# Create individual plots
plot_e0neg <- ggplot(results_e0neg, aes(x = avg_similarity, y = ext_prob, color = parameter)) +
  geom_line() +
  labs(title = "Effect of e_0neg on Extinction Probability", x = "Average Similarity", y = "Extinction Probability") +
  theme_minimal()

plot_e1neg <- ggplot(results_e1neg, aes(x = avg_similarity, y = ext_prob, color = parameter)) +
  geom_line() +
  labs(title = "Effect of e_1neg on Extinction Probability", x = "Average Similarity", y = "Extinction Probability") +
  theme_minimal()

plot_aeneg <- ggplot(results_aeneg, aes(x = avg_similarity, y = ext_prob, color = parameter)) +
  geom_line() +
  labs(title = "Effect of a_eneg on Extinction Probability", x = "Average Similarity", y = "Extinction Probability") +
  theme_minimal()

# Arrange the plots in a single row with separate legends
grid.arrange(plot_e0neg, plot_e1neg, plot_aeneg, ncol = 1)



```

**Effect of e_0neg (Base Extinction Probability):**

As e_0neg increases, the baseline extinction probability shifts upward. This directly increases the extinction probability for all levels of average similarity, serving as a starting point.

**Effect of e_1neg (Modifier for Extinction Probability):**

When e_1neg is negative, it decreases the extinction probability as similarity increases. Conversely, a positive e_1neg adds to the extinction probability, causing it to increase more rapidly with similarity.

**Effect of a_eneg (Decay Rate for Exponential Term):**

A higher a_eneg value makes the extinction probability decay more rapidly as average similarity increases. Lower values of a_eneg cause a slower decay, maintaining a higher probability over a more extensive range of similarity values.

```{r}
# Adjusted parameter values
pars$e_0neg <- 0.1    # Increase base extinction probability
pars$e_1neg <- -e_0neg    # Set a positive value to add to the extinction probability
pars$a_eneg <- 0.5    # Increase the rate of decay
```



# Probabilitstic niche

```{r}
# Load necessary libraries
library(ggplot2)

# Define the Gaussian probability function for interaction
prob_interaction <- function(distance, sigma) {
  exp(- (distance^2) / (2 * sigma^2))
}

# Create a range of distances from the optimal niche
distances <- seq(0, 1, length.out = 100)

# Define different sigma values to illustrate their effects
sigma_values <- c(0.05, 0.1, 0.2)

# Create a data frame to store the results
results <- data.frame()

# Compute the probabilities for each sigma value
for (sigma in sigma_values) {
  probabilities <- prob_interaction(distances, sigma)
  results <- rbind(results, data.frame(distance = distances, probability = probabilities, sigma = paste("sigma =", sigma)))
}

# Plot the probability of interaction for each sigma value using ggplot2
ggplot(results, aes(x = distance, y = probability, color = sigma)) +
  geom_line(size = 1) +
  labs(title = "Probability of Interaction Based on Distance from Optimal Niche",
       x = "Distance from Optimal Niche",
       y = "Interaction Probability") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("red", "blue", "green")) +
  theme(legend.position = "right")

```



# average similarity and extinction probability

```{r}
# Define parameters
e_0neg <- 0.1  # Base extinction probability
#e_1neg <- 0.2  # Modifier for extinction probability
e_1neg <- -e_0neg
a_eneg <- 0.5  # Decay rate

# Simulate data for average similarity from 0 to around 1.5
avg_similarity <- seq(0, 1.5, length.out = 100)

# Calculate extinction probabilities based on avg_similarity
ext_prob_sel <- e_0neg + e_1neg * exp(-a_eneg * avg_similarity)

# Plotting
plot(avg_similarity, ext_prob_sel, type = "l", col = "blue", lwd = 2,
     xlab = "Average Similarity", ylab = "Extinction Probability",
     main = "Extinction Probability vs. Average Similarity")
grid()

```

```{r}
# Define parameters
e_0neg <- 0.1  # Base extinction probability
e_1neg <- -e_0neg  # Ensure positive relationship between similarity and extinction probability
a_eneg <- 0.5  # Decay rate

# Simulate data for average similarity from 0 to around 1.5
avg_similarity <- seq(0, 0.5, length.out = 100)

# Rescale avg_similarity to a range between 0 and 1
avg_similarity_scaled <- (avg_similarity - min(avg_similarity)) / (max(avg_similarity) - min(avg_similarity))

# Set parameters for the logistic function
k <- 10  # Steepness of the logistic curve
midpoint <- 0.5  # Midpoint for the logistic curve (can be adjusted based on the desired mean similarity)

# Calculate extinction probabilities using a logistic function
ext_prob_sel <- 1 / (1 + exp(-k * (avg_similarity_scaled - midpoint)))

# Plotting
plot(avg_similarity, ext_prob_sel, type = "l", col = "blue", lwd = 2,
     xlab = "Average Similarity", ylab = "Extinction Probability",
     main = "Extinction Probability vs. Average Similarity using Logistic Function")
grid()

```

# Possible range of average similarities in different network sizes

```{r}
# Load necessary libraries
library(ggplot2)

# Function to calculate average similarity using SVD
calculate_avg_similarity <- function(L) {
  svd_result <- svd(L)
  similarity_matrix <- svd_result$v %*% diag(svd_result$d) %*% t(svd_result$v)
  diag(similarity_matrix) <- 0  # Set the diagonal to 0 to exclude self-similarity
  avg_similarity <- rowMeans(similarity_matrix) * (ncol(similarity_matrix) / (ncol(similarity_matrix) - 1))
  avg_similarity[avg_similarity < exp(-5)] <- 0  # Set small values to 0
  return(mean(avg_similarity, na.rm = TRUE))
}

# Simulate networks of sizes ranging from 2 to 70 species
network_sizes <- 2:70
avg_similarities <- data.frame(Size = integer(), Avg_Similarity = numeric())

# Simulate multiple networks for each size
set.seed(42)  # For reproducibility
for (size in network_sizes) {
  for (i in 1:100) {  # Simulate 100 networks for each size
    # Generate a random binary interaction matrix (0 or 1)
    L <- matrix(sample(c(0, 1), size * size, replace = TRUE), nrow = size)
    
    # Ensure no self-interactions by setting the diagonal to 0
    diag(L) <- 0
    
    # Calculate the average similarity
    avg_similarity <- calculate_avg_similarity(L)
    
    # Store the result in the data frame
    avg_similarities <- rbind(avg_similarities, data.frame(Size = size, Avg_Similarity = avg_similarity))
  }
}

# Plotting with ggplot2
ggplot(avg_similarities, aes(x = factor(Size), y = Avg_Similarity)) +
  geom_boxplot(outlier.size = 0.5, fill = 'skyblue', color = 'darkblue') +
  labs(title = "Distribution of Average Similarities by Network Size",
       x = "Network Size (Number of Species)",
       y = "Average Similarity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}

# Load necessary library
library(ggplot2)

# Function to calculate establishment probability
calculate_estab_prob <- function(sum_I, u_0pos, u_1pos, a_upos) {
  estab_prob_sel <- u_0pos + u_1pos * (1 - exp(-a_upos * sum_I))
  return(estab_prob_sel)
}

# Define a range of integer values for sum_I
sum_I_values <- 1:10  # Integer range from 1 to 10

# Parameters to test
u_0pos <- 0.1   # Baseline probability
u_1pos <- 0.5   # Scaling factor
a_upos <- 0.5   # Shape of the exponential increase

# Calculate establishment probabilities for different sum_I values
estab_prob_values <- sapply(sum_I_values, calculate_estab_prob, u_0pos, u_1pos, a_upos)

# Create a data frame for plotting
plot_data <- data.frame(sum_I = sum_I_values, estab_prob = estab_prob_values)

# Plot the result
ggplot(plot_data, aes(x = sum_I, y = estab_prob)) +
  geom_line(color = 'blue') +
  geom_point(size = 2, color = 'red') +
  labs(title = "Establishment Probability vs. Sum of Interactions",
       x = "Sum of Interactions (sum_I)",
       y = "Establishment Probability (estab_prob_sel)") +
  theme_minimal()

# Create a Shiny app to play with parameter values
if (interactive()) {
  library(shiny)
  
  ui <- fluidPage(
    titlePanel("Test Different Parameter Values for Establishment Probability"),
    sidebarLayout(
      sidebarPanel(
        sliderInput("u_0pos", "Baseline Probability (u_0pos):", min = 0, max = 1, value = 0.1, step = 0.01),
        sliderInput("u_1pos", "Scaling Factor (u_1pos):", min = 0, max = 1, value = 0.5, step = 0.01),
        sliderInput("a_upos", "Exponential Shape (a_upos):", min = 0, max = 2, value = 0.5, step = 0.01)
      ),
      mainPanel(
        plotOutput("estabPlot")
      )
    )
  )
  
  server <- function(input, output) {
    output$estabPlot <- renderPlot({
      # Recalculate establishment probabilities with the input parameters
      estab_prob_values <- sapply(sum_I_values, calculate_estab_prob, input$u_0pos, input$u_1pos, input$a_upos)
      
      # Create a data frame for plotting
      plot_data <- data.frame(sum_I = sum_I_values, estab_prob = estab_prob_values)
      
      # Plot the result
      ggplot(plot_data, aes(x = sum_I, y = estab_prob)) +
        geom_line(color = 'blue') +
        geom_point(size = 2, color = 'red') +
        labs(title = "Establishment Probability vs. Sum of Interactions",
             x = "Sum of Interactions (sum_I)",
             y = "Establishment Probability (estab_prob_sel)") +
        theme_minimal()
    })
  }
  
  shinyApp(ui = ui, server = server)
}


```

```{r}
# Load necessary library
library(ggplot2)

pars$u_0pos <- 0.3   # Baseline probability
pars$u_1pos <- 0.7  # Scaling factor
pars$a_upos <- 1   # Shape of the exponential increase

# Custom function for establishment probability
custom_estab_prob <- function(sum_I, u_0pos = 0.3, u_1pos = 0.7, a_upos = 1) {
  # Use an exponential function with an asymptote
  # Ensure the probability starts at base_prob for sum_I = 1 and reaches max_prob at sum_I = 10
  prob <- u_0pos + (u_1pos - u_0pos) * (1 - exp(-a_upos * (sum_I - 1)))
  return(prob)
}

# Define a range of integer values for sum_I
sum_I_values <- 1:10  # Integer range from 1 to 10

# Calculate establishment probabilities using the custom function
estab_prob_values <- sapply(sum_I_values, custom_estab_prob, u_0pos = 0.3, u_1pos = 0.7, a_upos = 0.5)

# Create a data frame for plotting
plot_data <- data.frame(sum_I = sum_I_values, estab_prob = estab_prob_values)

# Plot the result
ggplot(plot_data, aes(x = sum_I, y = estab_prob)) +
  geom_line(color = 'blue') +
  geom_point(size = 2, color = 'red') +
  labs(title = "Establishment Probability vs. Sum of Interactions",
       x = "Sum of Interactions (sum_I)",
       y = "Establishment Probability") +
  theme_minimal()

```

