---
title: "Phylogenetic signal in Madelaine's Galapagos data"
output: html_notebook
---

```{r}
library("igraph")
library("bipartite")
library("ape")
library("RcppEigen")
library('inline')
library('magrittr')
library("phylolm")
library("ade4")
library("FactoMineR")
library("vegan")
library("dummies")
library("rnetcarto")
library("NetIndices")
library("phylobase")
library("cluster")

library("funrar")
library(ggplot2)

```

```{r}
interact_databade <- read.csv("../Phylo.sig_Gal/Data/Interactions.csv",sep=";", header = TRUE)
cleared.metaweb<-readRDS("../Phylo.sig_Gal/Data/Metaweb_pariles_vertebrate_Genus.RDS", refhook = NULL)
sp_list<-read.csv("../Phylo.sig_Gal/Data/sp_list.csv",sep=";", header = TRUE)
cleared.interact_db<-readRDS("../Phylo.sig_Gal/Data/Inter_predproie_extrapoles_Genus_invertebresgenus_07122020.RDS", refhook = NULL)
```

## Obtain adjacency matrix
```{r}
# Convert the data into a graph object using the first and third columns of the dataset as an edgelist
graph<-graph.edgelist(as.matrix(cleared.interact_db[,c(1,3)]), directed=TRUE)
n=vcount(graph)
# Get Adjacency matrix
Adj = get.adjacency(graph)
names = row.names(Adj)

plot_graph<-plot.igraph(graph,vertex.label=NA,vertex.size=3,edge.arrow.size=.25)
```


## Decomposition of Adjacency matrix in singular values to obtain traits of interest
```{r}
#compute Singular Value Decomposition
SVD = svd(Adj)
U = SVD$u
S = diag(SVD$d)
Ssqrt = structure(vapply(S, sqrt, numeric(1)),dim=dim(S))
V = SVD$v

#Obtain traits
#compute full rank in- and out- traits
traits_in =  V %*% Ssqrt
traits_out = U %*% Ssqrt
row.names(traits_in) = row.names(traits_out) = names
traits_in_df = data.frame(traits_in)
traits_out_df = data.frame(traits_out)

ncol(traits_out_df)
```


## Determine number of dimensions
```{r}
#choix du nombre de vecteurs singuliers (=traits de pollinisation) a prendre en compte dans la RDA
singular.values<-SVD$d
#plot de la somme cumulee de la proportion d'inertie expliquee par les traits
plot(cumsum(singular.values/sum(singular.values))) # ~90
#from Dalla Riva
getElbows(SVD$d[SVD$d > quantile(SVD$d,0.05)]) #-> elbows:  4, 27, 80
```




# Obtaining the phylogenetic distance matrix

```{r}
# Phylogenetic tree
phylo.tree <- phylobase::readNexus("../Phylo.sig_Gal/Data/GPS_all_verts.nwk")
phylo.tree <- as(phylo.tree, "phylo")

#Corriger les erreurs d'ortographe dans phylo.tree$tip
phylo.tree$tip.label <- stringi::stri_replace_all(phylo.tree$tip.label, replacement = " ", regex = "_")
phylo.tree$tip.label[which(phylo.tree$tip.label == "Lasirurs cinereus")] <- "Lasiurus cinereus"
phylo.tree$tip.label[which(phylo.tree$tip.label == "Microlophus grayi")] <- "Microlophus grayii"
phylo.tree$tip.label[which(phylo.tree$tip.label == "Microlophis delanonis")] <- "Microlophus delanonis"
phylo.tree$tip.label[which(phylo.tree$tip.label == "Microlophis bivittatus")] <- "Microlophus bivittatus"
phylo.tree$tip.label[which(phylo.tree$tip.label == "Microlophis habeli")] <- "Microlophus habelii"
# Drop absent species
dropme <- phylo.tree$tip.label[!phylo.tree$tip.label %in% sp_list$x]
sub.tree <- drop.tip(phylo.tree, dropme)
# Calcul de la matrice de distance et la mettre en ordre alphabétique
#dist_mat <- cophenetic(sub.tree)
#dist_mat <- dist_mat[order(rownames(dist_mat)),order(colnames(dist_mat))]
```


```{r}
#creation de la matrice de variance covariance phylogenetique a partir de l'arbre phylogenetique corrige
#orchid.corr.vcv<-vcv(orchid.tree.corr)
phy.corr.vcv<-vcv(sub.tree)
str(phy.corr.vcv)
#corr.phy_matrix<-cov2cor(phy.corr.vcv)

#Madelaine's method
dist_mat_phy_mad <- cophenetic(sub.tree)
dist_mat_phy_mad1 <- dist_mat_phy_mad[order(rownames(dist_mat_phy_mad)),order(colnames(dist_mat_phy_mad))]
```



## Number of singular vectors of phylogenetic correlation matrix

```{r}
#choix du nombre de vecteurs singuliers de la matrice de correlation phylogenetique a prendre en compte dans la RDA
sv_phylo<-svd(phy.corr.vcv)$d
plot(cumsum(sv_phylo/sum(sv_phylo))) # ~22


#with matrix obtained with madelaine's method
sv_phylo1<-svd(dist_mat_phy_mad1)$d
plot(cumsum(sv_phylo1/sum(sv_phylo)))
```




# Loop to analyse how correlation values respond to dimensionality
```{r}

############# Select only those species used in phylogenetic analyses
matched_diets.matrix <- subset(traits_out_df , rownames(traits_out_df) %in% rownames(phy.corr.vcv))
View(matched_diets.matrix)





############# Order phylogenetic_distance dataframe to match rownames order of spp x dimensions (diets) dataframe.

df.phy_dist<-as.data.frame(phy.corr.vcv) #first convert to a dataframe

 
df.phy_dist.ord1 <- df.phy_dist[match(row.names(matched_diets.matrix),row.names(df.phy_dist)),] #second, we order the rownames as in spp x dimensions matrix (diets matrix).

mat_ord <- function(mx) mx[, c(rownames(mx), setdiff(colnames(mx), rownames(mx)))] #We match the order of columns with the new rows' order
df.phy_dist.ord2<-mat_ord(df.phy_dist.ord1)







############# these are the dataframes I will convert into distance matrices and compare
View(matched_diets.matrix)
View(df.phy_dist.ord2)





############# LOOP to create all the mdiet matrices with certain d 


#1) create all dataframes with increasing number of d

list.df.diets_d<-list()
for(i in 1:ncol(matched_diets.matrix)) {       # for-loop over columns
  list.df.diets_d[[i]] <- matched_diets.matrix[ , 1:i,drop=FALSE]
}
class(list.df.diets_d)
View(list.df.diets_d)

#Check that they are correctly ordered (coinciding with rownames of dist.matrix.phy)
View(list.df.diets_d[[23]])
View(df.phy_dist.ord2) #these are well matched


#2) convert to distance matrices
dist.matrix.phy<- daisy(df.phy_dist.ord2, metric="euclidean")

list.dist.diets_d<-lapply(list.df.diets_d, FUN = daisy)


# 3) Save distance objects as matrices
list_matrix.distance_diets<-lapply(list.dist.diets_d, FUN = as.matrix)
matrix.distance_phy<-as.matrix(dist.matrix.phy)






############# Procrustes test 

#trial with one pair
protest.trial<-protest(list_matrix.distance_diets[[23]],matrix.distance_phy)
str(protest.trial)


list.corr.protest<-list()
list.corr.protest_sign<-list()
for(i in 1:length(list_matrix.distance_diets)) {       # for-loop over columns
  #list.corr.protest[[i]] <- protest(list_matrix.distance_diets[[i]],matrix.distance_phy)$t0
  list.corr.protest_sign[[i]] <- protest(list_matrix.distance_diets[[i]],matrix.distance_phy)$signif
}

View(list.corr.protest)
View(list.corr.protest_sign)

# Convert the lists to dataframes
corr.vector<-unlist(list.corr.protest)
corr.df<-as.data.frame(corr.vector)

signif.vector<-unlist(list.corr.protest_sign)
signif.df<-as.data.frame(signif.vector)

# Create column d (dimensions)
corr.df$d <- 1:nrow(corr.df)
signif.df$d <- 1:nrow(signif.df)
#unlist(list.corr.protest)


#Merge both dataframes by d
Dataframe_d_corr<-merge(corr.df, signif.df, by = "d")


ggplot(Dataframe_d_corr, aes(x = d, y = corr.vector)) + 
  geom_point() + geom_line(aes(colour="red")) 
```

Correlation in just 5 dimensons is 27%. In less than 10 (9), 30%. Increasing the dimensons up to 120, only increases the correlation until the maximum 36%. 
We observe decreases in correlation with the addition of some dimensions (at around 70 and 125d).







#######################################################



## phylog. signal of predator diets

```{r}
matrix.diets.y<-traits_in_df[,1:90]
matched_diets.matrix <- subset(matrix.diets.y, rownames(matrix.diets.y) %in% rownames(phy.corr.vcv))
View(matched_diets.matrix)
View(corr.phy_matrix)

dist.matrix.diets<- compute_dist_matrix(matched_diets.matrix, metric="euclidean")
View(dist.matrix.diets)

dist.matrix.phy<- compute_dist_matrix(phy.corr.vcv, metric="euclidean")
View(dist.matrix.phy)

Matrice.dist.diets <- dist.matrix.diets[order(rownames(dist.matrix.diets)), order(colnames(dist.matrix.diets))]

Matrice.dist.phy <- dist.matrix.phy[order(rownames(dist.matrix.phy)), order(colnames(dist.matrix.phy))]

View(Matrice.dist.diets)
View(Matrice.dist.diets)

mantel.test(Matrice.dist.diets,Matrice.dist.phy) #p=0.067
mantel.test(Matrice.dist.diets,dist_mat_phy_mad1) #0.459
```

```{r}
matched_diets.matrix <- subset(Dist_1d.m , rownames(Dist_1d) %in% rownames(phy.corr.vcv))
View(matched_diets.matrix)
View(corr.phy_matrix)
```


procrustes
```{r}
pcoa_pred.j <- wcmdscale(Matrice.dist.diets, eig = T, add = "cailliez") 
pcoa.phylo <- wcmdscale(Matrice.dist.phy, eig = T, add = "cailliez")

procustes.fit <- protest(pcoa_pred.j,pcoa.phylo)
procustes.fit

plot(procustes.fit, type="text", kind=1, cex=0.1)


data(varespec)
vare.dist <- vegdist(wisconsin(varespec))
mds.null <- monoMDS(vare.dist, y = cmdscale(vare.dist))
mds.alt <- monoMDS(vare.dist)
vare.proc <- procrustes(mds.alt, mds.null)
vare.proc
summary(vare.proc)

```



same but considering all dimensions, not only 90
```{r}
matched_diets.matrix2 <- subset(traits_in_df, rownames(traits_in_df) %in% rownames(phy.corr.vcv))


dist.matrix.diets2<- compute_dist_matrix(matched_diets.matrix2, metric="euclidean")


dist.matrix.phy<- compute_dist_matrix(phy.corr.vcv, metric="euclidean")
View(dist.matrix.phy)

Matrice.dist.diets2 <- dist.matrix.diets2[order(rownames(dist.matrix.diets2)), order(colnames(dist.matrix.diets2))]

Matrice.dist.phy <- dist.matrix.phy[order(rownames(dist.matrix.phy)), order(colnames(dist.matrix.phy))]

View(Matrice.dist.diets)
View(Matrice.dist.diets)

mantel.test(Matrice.dist.diets2,Matrice.dist.phy) #P=0.087
```
















