---
title: "R Notebook"

---

```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
library(ggpubr)
library(igraph)
library(EnvStats)
library(aricode)
library(vegan)
library(ade4)
library(ape)
library(tidyr)
library(ggraph)
```


Load results from 1 simulation (change file path as needed)

```{r}
res_sim <- readRDS(here::here("results_toplot_networks_timesteps.rds"))
```


#### Inspect simulation dynamics 

Load functions (change file path as needed)

```{r}
source(here::here("functions_simulation_inspection.R"))
inspect_simulation_fw(simulation_data = res_sim, nbasals = pars$Sbasal, Smax = pars$Smax)
```





## Function to compute distance matrices

Taking as input the results from the simulation, this function does the following:

1) Prepares data
- Identifies and removes timesteps where phylogenetic distances can't be computed.
- Converts species identifiers from numbers to letters for consistency.

2) Processes phylogenetic data:

- Constructs phylogenetic distance matrices for living species at each timestep.
- Ensures species consistency between phylogenetic and presence matrices.
- Egeinvectors of the cropped phylogenetic correlation matrix are computed using:
list_svd_eigen.phy[[i]] <- eigen(list_phylo.corr_cropped[[i]], symmetric = T)$vec

3) Processes interaction Data:

- Filters interaction matrices to include only present species.
- Performs Singular Value Decomposition (SVD) on interaction matrices and keeps the right singular vectors (predator roles of species). The transposed right singular vectors are kept.


It saves the computed eigenvectors from the phylogenetic correlation matricesand the ones from the interaction matrices. We will use this output to compute correlations between them.


```{r}
compute_dist_matrices <- function(results_simulation, int, Smax, nbasals) {
  
  presence_matrix <- results_simulation$presence_matrix
  
  # number of timesteps
  n_steps <- length(results_simulation$network_list)
  
  #### Identify timesteps where phylogenetic distances cant be calculated
  non.valid_timesteps_phylo_distance <- c(which(rowSums(presence_matrix) < 3))
  # until what timestep need to discard:
  final.discarded_timestep <- (non.valid_timesteps_phylo_distance[length(non.valid_timesteps_phylo_distance)])+1
  
  #### homogenize elements to start from valid timesteps
  # ancestry table
  list_anc_dist <- results_simulation$list_anc_dist[(final.discarded_timestep+1):length(results_simulation$list_anc_dist)]
  # network list
  network_list <- results_simulation$network_list[(final.discarded_timestep+1):length(results_simulation$list_anc_dist)]
  
  
  ## ------------------- Check
  if (length(which(is.null(results_simulation$network_list))) > 0) {
    print("PROBLEM - null network somewhere")
  } 
  if(length(list_anc_dist) != length(network_list)){
    print("PROBLEM - length list_anc_dist != length(network_list)")
  }
  ## -------------------
  
  
  
  if(int == "foodweb"){
    #### Eliminate basal species
    Sbasals <- nbasals
    network_list <- lapply(network_list, eliminate_basals, nbasals = Sbasals)
  }
  
  #### Convert spp names from numbers to letters
  ## ancestry-distances table
  list_anc_dist_letters <- lapply(list_anc_dist, change_sppnames_letters_ancdist.table)
  ## Network list 
  list_networks_sppnames_numbers <- lapply(network_list, set_sppNames_numbers)
  list_networks_sppnames_letters <- lapply(list_networks_sppnames_numbers, convert_sppnames_toletters)
  
  
  
  ## ------------------- Check
  if(length(list_anc_dist_letters) != length(list_networks_sppnames_letters)){
    print("PROBLEM - list_anc_dist_letters != list_networks_sppnames_letters")
  } 
  ## -------------------
  
  
  #### convert spp names to letters in presence matrix
  colnames(presence_matrix) <- seq(1:Smax)
  colnames(presence_matrix) <- chartr("0123456789", "ABCDEFGHIJ", colnames(presence_matrix))
  
  
  
  ## ------------------- Check
  
  if(length(which(colnames(presence_matrix) != colnames(list_networks_sppnames_letters[[1]]))) != 0){
    print("PROBLEM - spp names in presence_matrix dont correspond to spp names in list_networks_sppnames_letterst")
  }
  ## -------------------
  
  
  
  # Discard same timesteps (rows) than the discarted phylogenetic distance matrices
  presence_matrix <- presence_matrix[(final.discarded_timestep+1):length(results_simulation$list_anc_dist),]
  
  
  ## ------------------- Check
  
  if(nrow(presence_matrix) != length(list_networks_sppnames_letters)){
    print("PROBLEM - presence matrix, list phylo dist and list interaction networks dont have the same n_steps")
  }
  
  ## -------------------
  
  ## Loop for obtaining phylogenetic distances:
  list_svd_eigen.phy <- list()
  list_phylo.corr_cropped <- list()
  
  
  for (i in 1:length(list_anc_dist_letters)) {
    
    print(paste("step", i))
    
    # obtain phylogenetic distances between all tree nodes
    
    newick <- ToPhylo2(list_anc_dist_letters[[i]])
    newick_tail <- paste(newick, "root")
    tree <- read.tree(text = sub("A root",";",newick_tail))
    #list_dist.phylo[[i]] <- cophenetic.phylo(tree)
    
    tree$edge.length<-sapply(tree$edge.length,function(x) ifelse(x==0,1e-5,x))
    phylo.vcv<-vcv(tree)
    phylo.corr<-cov2cor(phylo.vcv)
    
    # Crop matrix of phylogenetic distances with those species that are alive
    
    # alive_species <- list_anc_dist_letters[[i]][which(list_anc_dist_letters[[i]]$`A/E` == "A"), "spp"]
    alive_species <- names(which(presence_matrix[i, ] == 1))
    list_phylo.corr_cropped[[i]] <- phylo.corr[alive_species, alive_species]
    
    # Check that species in the phylogenetic distance matrix are the same present in the presence matrix
    
    if (all(names(which(presence_matrix[i, ] == 1)) != colnames(list_phylo.corr_cropped[[i]]))) {
      cat("Species mismatch detected for index", i, "\n")
    }
    
    
    list_svd_eigen.phy[[i]] <-eigen(list_phylo.corr_cropped[[i]], symmetric = T)$vec
  }
  
  
  ## ------------------- Check
  if (length(which(unlist(lapply(list_svd_eigen.phy, is.null)) == TRUE)) != 0) {
    print("PROBLEM - nulls in list_dist.phylo")
  }
  if (length(list_svd_eigen.phy) != length(list_networks_sppnames_letters)){
    print("PROBLEM - length(list_dist.phylo) != length(list_networks_sppnames_letters)")
  }
  ## -------------------
  
  
  
  ## ------------------- Check
  #Check that phylogenetic distance matrices retain present species:
  vec_error <- c()
  for (i in 1:nrow(presence_matrix)) {
    vec_tf <-  names(presence_matrix[i, which(presence_matrix[i,] == 1)]) == colnames(list_svd_eigen.phy[[i]])
    if(length(which(vec_tf == FALSE)) > 0){
      vec_error[i] <- "error"
    } else if (length(which(vec_tf == FALSE)) == 0 ){
      vec_error[i] <- "g"
    }
  }
  if(length(which(vec_error == "error")) > 0){
    print("PROBLEM - list_dist.phylo_pres dont retain present species")
  }
  ## -------------------
  
  
  #### Retain only present species in network matrices
  list_net_present_spp.letters <- list()
  
  for (i in 1:length(list_networks_sppnames_letters)) {
    list_net_present_spp.letters[[i]] <- list_networks_sppnames_letters[[i]][names(which(presence_matrix[i,] == 1)), names(which(presence_matrix[i,] == 1))]
  }
  
  
  
  
  
  #############################################################################
  
  
  list_svd_pred <- list()
  
  for (i in 1:length(list_net_present_spp.letters)) {
    
    # Perform SVD
    svd_result <- svd(list_net_present_spp.letters[[i]])
    
    # Extract left (U), middle (D), and right (V) matrices
    U <- svd_result$u       # Left matrix (n_prey x kept axes)
    D <- diag(svd_result$d) # Middle matrix (kept axes x kept axes)
    V <- svd_result$v       # Right matrix (n_predators x kept axes)
    
    # Decide on the number of axes to keep
    kept_axes <- ncol(list_net_present_spp.letters[[i]])  
    
    # Select the kept axes from U, D, and V
    U_kept <- U[, 1:kept_axes]   # n_prey x kept axes
    D_kept <- D[1:kept_axes, 1:kept_axes] # kept axes x kept axes
    V_kept <- V[, 1:kept_axes]   # n_predators x kept axes
    
    # Transpose the right matrix V to match the desired output
    list_svd_pred[[i]] <- t(V_kept)  
    
  }
  
  
  
  ## ------------------- Check
  if(length(list_svd_pred) != length(list_svd_eigen.phy)){
    print("PROBLEM - length(list_interact_distances_mean_corrected) != length(list_dist.phylo_pres")
  }
  if(length(list_svd_eigen.phy) != length(list_svd_pred)) {
    print("PROBLEM - list phylo dist and lists interact dist dont have the same length")
  } 
  
  vec_problems_ncol <- c()
  
  for (i in 1:length(list_svd_pred)) {
    if(ncol(list_svd_pred[[i]]) != ncol(list_svd_eigen.phy[[i]])){
      vec_problems_ncol[i] <- "P"
    } else if(ncol(list_svd_pred[[i]]) == ncol(list_svd_eigen.phy[[i]])){
      vec_problems_ncol[i] <- "g"
    }
  }
  if(length(which(vec_problems_ncol == "P") > 0)){
    print("PROBLEM - Interact and phylo dist. matrices dont have the same ncols")
  }
  
  vec_problems_sppcomp <- c()
  
  for (i in 1:length(list_svd_eigen.phy)) {
    vec_problems_sppcomp[i] <- identical(sort(colnames(list_svd_pred[[i]])), sort(colnames(list_svd_eigen.phy[[i]])))
  }
  if(length(which(vec_problems_ncol == "FALSE") > 0)){
    print("PROBLEM -  Interact and phylo dist. matrices dont have the order of colnames")
  }
  
  vec_problems <- c()
  
  for (i in 1:length(list_svd_pred)) {
    vec_truefalse <- colnames(list_svd_pred[[i]]) == colnames(list_svd_eigen.phy[[i]])
    if(FALSE %in% vec_truefalse){
      vec_problems[i] <- "P"
    }else{
      vec_problems[i] <- "_"
    }
  }
  if(length(which(vec_problems == "P") > 0)){
    print("PROBLEM - Interact and phylo dist. matrices dont have the same order of colnames")
  }
  
  
  
  result <- list("list_svd_pred" = list_svd_pred,
                 "list_svd_eigen.phy" = list_svd_eigen.phy,
                 "list_net_present_spp.letters" = list_net_present_spp.letters,
                 "list_phylo.corr_cropped" = list_phylo.corr_cropped)
  
  return(result)
}


```



Run function on results from 1 simulation:

```{r}
sdv_matrices <- compute_dist_matrices(results_simulation = res_sim,
    		                       int = "foodweb",
    		                      Smax = pars$Smax,
    		                      nbasals = pars$Sbasal)

list_svd_pred <- sdv_matrices$list_svd_pred
list_svd_eigen.phy <- sdv_matrices$list_svd_eigen.phy

#list_network <- sdv_matrices$list_net_present_spp.letters
#list_corrphylo <- sdv_matrices$list_phylo.corr_cropped
```



# Compute phylosignal


## Mantel corr between all axes of sdv


```{r}
# Initialize vectors to store results
timestep <- vector()
S <- vector()
cor_mantel <- vector()

# Loop through each timestep to compute Mantel correlations
for (i in seq_along(list_svd_pred)) {
  
  # Extract SVD results for prediction and phylogenetic matrices
  svd_pred <- list_svd_pred[[i]]
  svd_phy <- list_svd_eigen.phy[[i]]
  
  # Compute pairwise Euclidean distance matrices for both
   dist_pred <- dist(svd_pred)
   dist_phy <- dist(svd_phy)
  
  
  # Run Mantel test to compare the two distance matrices
  mantel_result <- mantel(dist_phy, dist_pred, permutations = 999)
  
  # Store results
  timestep[i] <- i
  S[i] <- ncol(list_svd_pred[[i]])  # Number of species (community size)
  cor_mantel[i] <- mantel_result$statistic  # Mantel correlation coefficient
}

df_mantel_results <- data.frame(timestep = timestep, 
                                S = S, 
                                cor_mantel = cor_mantel)

ggplot(df_mantel_results, aes(x = timestep, y = cor_mantel)) +
  geom_point(alpha = 0.5) +  
  labs(x = "Time Steps", y = "Mantel Correlation") +
  theme_classic() +
  geom_vline(xintercept = I_max, linetype = "dashed") +
  ylim(0, 1)
```


## Manel corr between optimized axes


```{r}
# Initialize vectors to store results
timestep <- vector()
S <- vector()
cor <- vector()
d_phylo <- vector()
d_network <- vector()
threshold <- 0.9  # Threshold for explained variance (e.g., 90%)

# Loop through each timestep to compute Procrustes correlations
for (i in seq_along(list_svd_pred)) {
  
  # Extract singular values from the original SVD for the interaction matrix
  svd_result <- svd(list_network[[i]])
  singular_values <- svd_result$d
  
  # Compute cumulative explained variance for the interaction matrix
  cum_var_pred <- cumsum(singular_values^2) / sum(singular_values^2)
  
  # Compute cumulative explained variance for the phylogenetic matrix
  eigenvalues_phy <- eigen(list_corrphylo[[i]], symmetric = TRUE)$values
  cum_var_phy <- cumsum(eigenvalues_phy) / sum(eigenvalues_phy)
  
  # Determine the number of axes needed to reach the threshold (90% variance)
  num_axes_pred <- min(which(cum_var_pred >= threshold))
  num_axes_phy <- min(which(cum_var_phy >= threshold))
  
  # Store the number of dimensions retained for each matrix
  d_network[i] <- num_axes_pred
  d_phylo[i] <- num_axes_phy
  
  # Decide on the number of axes to keep: the maximum of the two
  num_axes_to_keep <- max(num_axes_pred, num_axes_phy)
  
  # Adjust matrices to keep only the necessary axes
  svd_pred_kept <- list_svd_pred[[i]][, 1:num_axes_to_keep]
  svd_phy_kept <- list_svd_eigen.phy[[i]][, 1:num_axes_to_keep]
  
     dist_pred <- dist(svd_pred_kept)
   dist_phy <- dist(svd_phy_kept)
  
  # Run Procrustes analysis on the selected axes
  mantel_result <- mantel(dist_phy, dist_pred, permutations = 999)
  
   # Store results
  timestep[i] <- i
  S[i] <- ncol(list_svd_pred[[i]])  # Number of species (community size)
  cor_mantel[i] <- mantel_result$statistic  # Mantel correlation coefficient
}


df_mantel_results <- data.frame(timestep = timestep, 
                                S = S, 
                                cor_mantel = cor_mantel)

# Plot the Mantel correlation along time steps
ggplot(df_mantel_results, aes(x = timestep, y = cor_mantel)) +
  geom_point(alpha = 0.5) +  
  labs(x = "Time Steps", y = "Mantel Correlation") +
  theme_classic() +
  geom_vline(xintercept = I_max, linetype = "dashed") +
  ylim(0, 1)


```



Plot the dimensions retained for phylogenetic and network matrices:

```{r}

df_long <- reshape2::melt(df_results, id.vars = "timestep", measure.vars = c("d_phylo", "d_network"),
                          variable.name = "Dimension_Type", value.name = "Dimensions")

ggplot(df_long, aes(x = timestep, y = Dimensions, color = Dimension_Type)) +
  geom_point() + 
  geom_line() +  
  labs(x = "timestep", y = "Number of Dimensions Retained",
       color = "Dimension Type") +
  theme_classic() +
  scale_color_manual(values = c("d_phylo" = "blue", "d_network" = "red")) 
```







